Mathematical Language Processing \\ Project

### 2014-03-12 paper.tex
first complete draft of short paper

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{float}

@@

\usepackage{url}
\urldef{\mailsa}\path|rob@clabs.cc, schubotz@tu-berlin.de|
\newcommand{\todaydate}{\leadingzero{\day}.\leadingzero{\month}.\the\year}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\begin{abstract}
This paper presents automated definition discovery for
variables occurring in mathematical formulae. Ordinary natural language
processing approaches are trained based on large, manually annotated news
corpora. They fall short in reasonable quality of definition discovery for
scientific texts. In the \emph{Mathematical Language Processing} project, we
discover definitions
%we did not discover any relations  a relation would be something like "let
%<subj> be a <object>", or "we call <subj> <obj>", or "<subject> is <object>"
%or just "<subj> <obj>"
for identifiers in formulae from the surrounding text by statistical methods.
%with a machine learning approach for classification. This was also not done
%in this paper We extract the identifiers from the formulae and track co-
%occurrences of those in the surrounding text.
The evaluation of our prototypical system, applied on the Wikipedia text
corpus shows that our approach augments the user experience for articles with
mathematical formula substantially. While hovering the identifiers occurring
in the equation a pop-up with the most probable explanations for the
identifiers occurs. The displayed definition information provides a good match
to the actual meaning of the identifiers.
\keywords{machine learning, text mining, parallel computing}
\end{abstract}

@@

Mathematical formulae are a viable source of information for a wide range of
scientists. Often those formulas contain variables that are unknown or at
least ambiguous to the reader. Therefore, one needs to study the surrounding
text to find the relevant definition.

An automatic information retrieval system can be used to reduce the readers
effort to understand a formula. Especially students and scientists of other
disciplines would profit from such a system.

Unfortunately, no one has built a labeled text corpus that annotates variables
and their definition. Instead of labeling large amounts of text by hand, we
investigate a potential way of automatically labeling large corpora with a set
of statistical assumptions.

We chose the Wikipedia text corpus as a target because of two facts. First,
the mark-up for editing formulas is essentially a subset of \TeX. Therefore,
it enables us to collect all variables within the set of formulas in a
document. Second, we can leverage the semantic information between mark-up and
word tokens, e.g., hyperlinks.

The English Wikipedia contains roughly four million articles. Even if we only
pick articles containing \texttt{<math/>} mark-up, our processor still needs
to compute tens of thousands of articles. Especially when using a maximum
entropy \emph{POS} tagger \cite{Rathna96}, like the one in Stanford's NLP
framework, one can make use of a parallel processing system to speed up
computation.

We implement  show how the proposed strategy can be implemented in the PACT
programming model \cite{Alexandrov2010}.

\paragraph{Related Work}
\citeauthor{Quoc2010}\cite{Quoc2010} proposed an approach for
relating whole formulas to sentences and their describing paragraphs.
\citeauthor{Yokoi}\cite{Yokoi} trained a emph{support vector machine} to extract
natural language descriptions for mathematical expressions.

@@

\section{Implementation}
The MLP processing system \cite{github} is basically a set of PACTs, where
data flows through. In contrast to other programming models like Map/Reduce,
PACTs can operate on multiple inputs. At some point of time, one needs an
ordered list of tokenized sentences and a list of identified variables, to
compute the ranking score. A \emph{CoGroup Contract} takes both lists and
computes the subsets, build upon distinct document keys. Finally, a
\emph{Reduce Contract} filters out results below a given threshold $\theta$.

\tikzset{actor/.style={
        rectangle,
        minimum size=6mm,
        very thick,
        draw=gray!50!black!50,
        top color=white,
        bottom color=gray!50!black!20
    },
    arrow/.style={
        -latex, thick, shorten <=2pt,shorten >=2pt
    }
}
\begin{figure}[H]
	\begin{tikzpicture}[node distance=5mm and 8mm]
		\node (Input) [align=center]{Wiki Dumps};
		\node (DocumentParser) [actor, right=of Input, align=center] {\emph{Map}\\\textbf{Parser}};
		\node (Candidates) [actor, right=of DocumentParser, align=center] {\emph{CoGroup}\\\textbf{Kernel}};
		\node (Sentence) [actor, above=of DocumentParser, align=center] {\emph{Map}\\\textbf{Tagger}};
		\node (Filter) [actor, right=of Candidates, align=center] {\emph{Reduce}\\\textbf{Filter}};
		\node (Output) [right=of Filter,align=center]{Raw\ Candidates};
		\draw[arrow] (Input)--(DocumentParser);
		\draw[arrow] (DocumentParser)--(Sentence);
		\draw[arrow] (DocumentParser)--(Candidates);
		\draw[arrow] (Sentence.east)--(Candidates.north);
		\draw[arrow] (Candidates)--(Filter);
		\draw[arrow] (Filter)--(Output);
	\end{tikzpicture}
\caption{Data flow of the PACT program}
\end{figure}

There are three computation heavy parts in the processor. First, the raw
Wikipedia document dumps need to be transformed into plain text. At the same
step, the formulas need to be evaluated and replaced by place-holder words. We
used the \emph{WikiText} framework of the Mylyn project with a custom builder
to transform the document into plain text. For analysing the \TeX \ mark-up in
the \texttt{<math/>} tags, \emph{SnuggleTeX} transforms the formulas into
their MathML representations, where variable names can be easily extracted.

The second computation heavy part is tokenization and tagging of the plain
text documents. We used the Stanford maximum entropy tagger with the \emph{WSJ
left3words} language model.

Finally, the rating for all nouns, contained in all candidate sentences, needs
to be calculated.

%\subsection{Experiments}
\section{Evaluation}
%\subsubsection{Variable Retrieval}
\label{vr}
Throughout our experiments we made some observations that had an impact on the
accuracy of retrieving the correct set of variables. First of all, people tend
to incorrectly use \TeX\ trying to create formulas. E.g.,
\texttt{\textbackslash text\{log\}} is more often used than the correct
operator \texttt{\textbackslash log}. Another problem is that sometimes people
use indices as a form of annotation, like $T_{before}$, $R_{A}$, and $R_{B}$.
These ambiguities lead to a increased number of false positives.

We took a very conservative approach and preprocessed all formulas.
\texttt{\textbackslash text\{\}} blocks along with plain subscriptions will be
removed before analysis. Moreover, we created a comprehensive blacklist to
improve the results further. Short words like `a' and `i', which are very
common, could be easily matched by our processor if a formula contains one of
those symbols. Additionally, we blacklist common mathematical operators,
constants, and functions.

@@

\section{Conclusion}
Our experiments showed that combining a POS tagger with numerical statistics
about the text surface, can lead to quality results. However, this approach is
only applicable under certain conditions. In situations where both parts of a
relation are unknown, other methods, especially supervised ones, are used to
bootstrap a language model. Unfortunately, we still need a labelled test
corpus to measure the performance of a classifier, trained with our generated
data.

Apart from the mentioned weighting parameters, the ranking's quality is
influenced by two further parameters. These are the assumptions about the
parameter $c$. Balancing all these parameters requires extensive knowledge
about the underlying language. These parameters might be specific for every
type of relation. Therefore, this strategy depends heavily on human input and
many iterations of tuning.

To further improve the robustness of the \emph{term frequency} measure, one
might cluster sets of scientific documents, based on their specific field of
research.

%460 identifiers
\begingroup
\let\clearpage\relax
\bibliography{mlp-papers}
\endgroup

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{float}
\usepackage{todonotes}

@@

\usepackage{url}
\urldef{\mailsa}\path|rob@clabs.cc,schubotz@tu-berlin.de |
\newcommand{\todaydate}{\leadingzero{\day}.\leadingzero{\month}.\the\year}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\begin{abstract}
In natural language words and phrases themselves imply the semantics.
Complimentary, the meaning of identifiers in mathematical formulae is undefined.
Thus scientist must study the context to decode the meaning.
The \emph{Mathematical Language Processing} project, aims to support that process and
presents automated identifier definition discovery.

Ordinary natural language processing approaches are trained on large, manually annotated news
corpora. They fall short in reasonable quality of definition discovery for scientific texts.
We use statistical methods to discover definitions in the surrounding text.

The evaluation of our prototypical system, applied on the Wikipedia text
corpus shows that our approach augments the user experience
substantially. While hovering the identifiers occurring
in the formula tool-tips with the most probable definitions
occur.
Tests with random samples show, that the displayed definitions provide
a good match with the actual meaning of the identifiers.   
\keywords{machine learning, text mining, parallel computing}
\end{abstract}

@@

Mathematical formulae are a viable source of information for a wide range of
scientists. Often, they contain variables that are unknown or ambiguous to the reader.
Therefore, one needs to study the surrounding text to find the relevant definition.

An automatic information retrieval system can be used to reduce the readers
effort. Especially students and scientists of other
disciplines would profit from a system that helps them to understand formulae.

To build such a system a labeled text corpus that annotates variables
and their definition is desirable. At the project start, such a corpus was not available.
%Unfortunately, no one has built . Instead of labeling large amounts of text by hand,
Consequently, we use simple statistical assumptions about co-occurrences between identifiers
and their potential definition. In combination with the Stratosphere system
that is capable of scalable data processing,  we are able to generate a large amount of 
definition relation candidates in a reasonable and a little implementation effort.
% the data processing system Stratosphere we are able to analyze large text corpora   and a system that  investigate a potential way of automatically labeling large corpora with a set
%of statistical assumptions.

We chose the Wikipedia text corpus as a target because of two facts.
First, it is easy to extract the formulas and identifiers from the text,
since the input language for formula texvc is very restrictive.
Second, we can leverage the semantic information between mark-up and
word tokens, e.g., hyperlinks.

The English Wikipedia contains roughly four million articles. Even if we only
pick articles containing \texttt{<math/>} tag, our processor still needs
to compute tens of thousands of articles. Especially when using a maximum
entropy \emph{POS} tagger \cite{Rathna96}, like the one in Stanford's NLP
framework, one can make use of a parallel processing system to speed up
computation.

Therefore we implement the proposed strategy with the Stratosphere system based
on the PACT programming model \cite{Alexandrov2010}.

\paragraph{Related Work}
\citeauthor{Quoc2010}\cite{Quoc2010} proposed an approach for
relating whole formulas to sentences and their describing paragraphs.
\citeauthor{Yokoi}\cite{Yokoi} trained a \emph{support vector machine} to extract
natural language descriptions for mathematical expressions.

@@

\paragraph{Implementation}
We implemented the MLP processing system \cite{github} as Stratosphere data-flow using JAVA which allows for scalable execution, application of complex higher order functions and easy integration of third party tool like Stanford NLP and the Mylyn framework.
%
%In contrast to MapReduce that supports only the second order functions map and reduce, we used the CoGroup function 
%
%\tikzset{actor/.style={
%        rectangle,
%        minimum size=6mm,
%        very thick,
%        draw=gray!50!black!50,
%        top color=white,
%        bottom color=gray!50!black!20
%    },
%    arrow/.style={
%        -latex, thick, shorten <=2pt,shorten >=2pt
%    }
%}
%\begin{figure}[H]
%	\begin{tikzpicture}[node distance=5mm and 8mm]
%		\node (Input) [align=center]{Wiki Dump};
%		\node (DocumentParser) [actor, right=of Input, align=center] {\emph{Map}\\\textbf{Parser}};
%		\node (Candidates) [actor, right=of DocumentParser, align=center] {\emph{CoGroup}\\\textbf{Kernel}};
%		\node (Sentence) [actor, above=of DocumentParser, align=center] {\emph{Map}\\\textbf{Tagger}};
%		\node (Filter) [actor, right=of Candidates, align=center] {\emph{Reduce}\\\textbf{Filter}};
%		\node (Output) [right=of Filter,align=center]{Raw\ Candidates};
%		\draw[arrow] (Input)--(DocumentParser);
%		\draw[arrow] (DocumentParser)--(Sentence);
%		\draw[arrow] (DocumentParser)--(Candidates);
%		\draw[arrow] (Sentence.east)--(Candidates.north);
%		\draw[arrow] (Candidates)--(Filter);
%		\draw[arrow] (Filter)--(Output);
%	\end{tikzpicture}
%\caption{Data flow of the Stratosphere program: We used an
%ordered list of tokenized sentences and a list of identified variables, to
%compute the ranking score. A \emph{CoGroup Contract} takes both lists and
%computes the subsets, build upon distinct document keys. Finally, a
%\emph{Reduce Contract} filters out results below a given threshold $\theta$.}
%\end{figure}
%
%There are three computation heavy parts in the processor. First, the raw
%Wikipedia document dumps need to be transformed into plain text. At the same
%step, the formulas need to be evaluated and replaced by place-holder words. We
%used the \emph{WikiText} framework of the Mylyn project with a custom builder
%to transform the document into plain text. For analyzing the \TeX \ mark-up in
%the \texttt{<math/>} tags, \emph{SnuggleTeX} transforms the formulas into
%their MathML representations, where variable names can be easily extracted.
%
%The second computation heavy part is tokenization and tagging of the plain
%text documents. We used the Stanford maximum entropy tagger with the \emph{WSJ
%left3words} language model.
%
%Finally, the rating for all nouns, contained in all candidate sentences, needs
%to be calculated.
%

%\subsection{Experiments}
\section{Evaluation}
We extracted 460 identifiers definition tuples in 550000 relations with a
high score ($>0.8$). The most common identifier definition tuples were 
number $n$ (1709), time $t$	(1480), mass $M$ (1042), radius $r$ (752), temperature	$T$	(666), angle $\theta$ (639), group $G$ (635).
%\todo{I don't trust this evaluation and I'm trying to rewrite that}
%%\subsubsection{Variable Retrieval}
%\label{vr}
%Throughout our experiments we made some observations that had an impact on the
%accuracy of retrieving the correct set of variables. First of all, people tend
%to incorrectly use \TeX\ trying to create formulas. E.g.,
%\texttt{\textbackslash text\{log\}} is more often used than the correct
%operator \texttt{\textbackslash log}. Another problem is that sometimes people
%use indices as a form of annotation, like $T_{before}$, $R_{A}$, and $R_{B}$.
%These ambiguities lead to a increased number of false positives.
%
%We took a very conservative approach and preprocessed all formulas.
%\texttt{\textbackslash text\{\}} blocks along with plain subscriptions will be
%removed before analysis. Moreover, we created a comprehensive blacklist to
%improve the results further. Short words like `a' and `i', which are very
%common, could be easily matched by our processor if a formula contains one of
%those symbols. Additionally, we blacklist common mathematical operators,
%constants, and functions.

@@

\section{Further work}
Our first intuition was to discover grammatical patters like \emph{let \texttt{variable} be \texttt{definition}} or were \emph{\texttt{variable} (indicates/stands for/ denotes)\texttt{definition}} based on the statistical findings. However, our impression is now that the performance gain would bi little.

Now, we are thinking to improve the distance measure between identifier and definition by considering the grammatical structure of the sentence. I.e the amount of distance added to scoring function should depend on the grammatical type of the relation between the words.

To improve the robustness of the \emph{term frequency} measure further, one
might cluster sets of scientific documents, based on their specific field of
research.

\section{Conclusion}
Our experiments showed that combining a POS tagger with numerical statistics
about the text surface, can lead to quality results. However, this approach is
only applicable under certain conditions. 
If there is only few data for some identifier our statistical approach tends to fail. In that situations other methods, especially supervised ones, are used to
bootstrap a language model. Unfortunately, we still need a labeled test
corpus to measure the performance of a classifier, trained with our generated
data.

Furthermore it would be nice to have a systematic approach for a wise choice of the ranking parameters.

\begingroup
\let\clearpage\relax
\bibliography{mlp-papers}
\endgroup
